Objective-First Design of Nanophotonic Waveguide Couplers

### 2011-10-10 presentations/ob1_wg_paper/paper.tex
Copy of draft sent to Jelena.

\begin{document}
\title{Automated Design of Nanophotonic Waveguide Couplers}
\author{Jesse Lu and Jelena Vu\v{c}kovi\'{c}}
\address{Stanford University, Stanford, California, USA.}
\email{jesselu@stanford.edu}

\maketitle
\begin{abstract}
We introduce an ``objective-first'' approach to the design of
    nanophotonic waveguide couplers.
We show that our method is
    computationally fast (20 minutes), 
    requires no trial-and-error,
    does not require a good starting design,
    can be applied to arbitrary waveguide modes,
    and generates devices with high coupling efficiencies (typically 95\%)
    and small footprints (1-4 square vacuum wavelengths).
\end{abstract}
% \ocis{230.7370, 130.3990.}

% \begin{thebibliography}{99}
% \bibitem{lipson} Almeida et al, Optics Letters \textbf{28}, 1302-1304 (2003).
% \end{thebibliography}
% Importance of problem.

@@

% Why is waveguide mode conversion important?
\subsection{The importance of waveguide mode conversion}
Optical mode conversion, 
    the efficient transfer of photons from one guided mode to another,
    is a fundamental requirement in nanophotonics.
Efficient conversion between waveguides modes
    is critical in the many cases, including:
    \begin{enumerate}
        \item Coupling to and from optical fiber\cite{}, 
        to communicate with the outside world.
    \item Coupling between various nanophotonic waveguides, 
        since different waveguides are best suited for different applications.
        For example, ridge waveguides seem ideal for low-loss transport\cite{},
            but other waveguides, 
            such as photonic crystal waveguides or slot waveguides,
            may be better suited for slow-light\cite{} 
            or energy-focusing devices\cite{}.
    \item Coupling between different materials such as
       passive, active\cite{}, and non-linear\cite{} materials. 
    \end{enumerate}
Although efficient waveguide coupling is essential in any nanophotonic system,
    practical methods to design couplers have not yet been developed.

% Why are current solutions insufficient?
\subsection{Approaches to designing waveguide couplers}
% Brute force
Of the design strategies currently available,
    brute-force parameter search is the most often employed 
    to-date because of its sheer simplicity\cite{}.
Although it may be suitable for tuning existing designs\cite{},
    the parameter space for most practical devices is simply too large
    for such a strategy to be tractable.

% Adiabatic mode conversion (large devices, symmetry breaking)
Adiabatic mode conversion strategies have been succesful
    for certain fiber-waveguide\cite{} and waveguide-waveguide\cite{} couplers,
    even if the resulting devices require large footprints.
However, adiabatic approaches cannot be used in many important cases 
    where the mode structure is more complex, such as
    coupling from ridge to some photonic crystal waveguides\cite{},
    or where the geometry of the problem is non-trivial, such as
    coupling in the out-of-plane direction, or
    coupling between modes of opposite symmetry.

% Local optimization
On the other hand, optimization methods based on local topological derivatives 
    seem very promising\cite{}
    in that they are both much faster than brute-force methods and
    more adaptable than adiabatic strategies.
However, these methods still carry a significant computational burden 
    in that every updated design must be simulated at least once.
Also, there is a significant burden for the user from whom
    a good initial design is usually required.

% How does your algorithm solve this problem?
% \subsection{Advantages of objective-first design}
In contrast, our method
\begin{itemize}
    \item does not employ brute-force parameter searches,
    \item does not require a good initial design,
    \item is computationally fast (no simulations required),
    \item generates couplers between seemingly arbitrary waveguide modes, and
    \item can generate these couplers in a very small footprint. 
\end{itemize}
We outline the general design strategy,
    as well as how it is applied to nanophotonic design, in the section below.

@@

The differences between Eqs.~\ref{eq:adj} and \ref{eq:ob1} are:
\begin{enumerate}
    \item in Eq.~\ref{eq:adj}, we attempt to decrease the design objective 
            (Eq.~\ref{eq:adj:obj}), 
        while in Eq.~\ref{eq:ob1}, the design objective is kept at zero 
            (Eq.~\ref{eq:ob1:con}); 
    \item in Eq.~\ref{eq:adj}, we always satisfy the underlying physics 
            (Eq.~\ref{eq:adj:con}),
        while in Eq.~\ref{eq:ob1}, physics is not satisfied,
            since the physics residual is generally non-zero
            (Eq.~\ref{eq:ob1:obj}).
\end{enumerate}

@@

The first practical implication of the objective-first approach 
    is that the number of independent variables is increased to include
    both $x$ and $p$.
In Eq.~\ref{eq:adj}, the constraint that physics must be satisfied, $g(x,p)=0$, 
    essentially forces $x$ to be a dependentent variable,
    since the choice of $p$ implicitly determines the value of $x$
    (there is generally a one-to-one mapping from $p$ to $x$).
In contrast, Eq.~\ref{eq:ob1} allows both $x$ and $p$ to vary independently,
    because the constraint, $f(x)=0$, is only a function of $x$.

Secondly, the amount of computation needed to enforce the constraint is
    drastically reduced in the objective-first approach.
This is because Eq.~\ref{eq:adj} requires a full solution of $g(x,p)=0$
    (i.e. a full simulation of the structure, $p$) to compute $x$.
In contrast, the constraint $f(x)=0$ in Eq.~\ref{eq:ob1} 
    can often be enforced so quickly (as shown below) that 
    future implementations may even produce designs in the same amount of time
    as required to simulate them!

Lastly, the objective-first approach eliminates the need for even a reasonable
    initial design.
Generally, methods based on Eq.~\ref{eq:adj} require an initial design which
    already provides some limited functionality
    (e.g. a coupler which already transfers 
    a non-zero amount of power to the desired output mode).
In constrast, methods based on Eq.~\ref{eq:ob1} perform just as well
    when started from a completely non-functional design 
    (e.g. a coupler which transfers no power into the desired output mode).

In conclusion, an objective-first approach results in a method 
    that is computationally fast (since it does not require simulation), and
    that can be applied to non-intuitive problems 
    where a functional starting design is not readily available.

% How is it applied to waveguide coupler design?
\subsection{Objective-first approach to waveguide coupler design}
We now apply an objective-first approach to the problem of designing 
    two-dimensional nanophotonic waveguide couplers.

We choose to work in the two-dimensional transverse electric mode,
    which only couples $E_x$, $E_y$, and $H_z$ ($E_z, H_x, H_y = 0$),
    since it is most relevant for on-chip devices\cite{}.
We choose to use $H_z$ as the field variable, 
    and $\epsilon^{-1}$ (inverse of the permittivity) 
    as the structure variable.
This results in the following representation of the physics residual, 
    based on the time-harmonic electromagnetic wave equation without sources;
    \begin{equation}
    \|g(H_z, \epsilon^{-1})\|^2 = 
    \| \nabla \times \epsilon^{-1} \nabla \times H_z - \mu \omega^2 H_z \|^2,
    \end{equation}
    where $\omega$ is the angular frequency,
    and $\mu$ is the permeability of free-space.

@@

Finally, in any objective-first approach,
    the physics residual is never guaranteed to decrease to zero.
Thus, it is entirely possible to never achieve 
    a physically realizable field, $H_z$.
In such cases, which are the norm rather than the exception,
    we find that a relatively small residual usually leads to
    fairly good, although imperfect, device performance.

\subsection{Numerical methods used to solve the objective-first design problem}
The design problem is now
    \begin{subequations}\label{eq:act}
    \begin{align} 
    \text{decrease} & \quad  
        \| \nabla \times \epsilon^{-1} \nabla \times H_z - \mu \omega^2 H_z \|^2
        \label{eq:act:obj} \\ 
    \text{subject to} & \quad 
        \left. H_z - H_z^\text{perfect} \right| _\text{boundary} = 0 
        \label{eq:act:con1} \\
        & \quad 
        \left. \frac{\partial H_z}{\partial n} - 
        \frac{\partial H_z^\text{perfect}}{\partial n} \right|_\text{boundary}
            = 0. \label{eq:act:con2}
    \end{align}
    \end{subequations}

This problem contains many local minima (it is non-convex\cite{});
    however, when either the field ($H_z$) or the structure ($\epsilon^{-1}$)
    variable is considered separately Eq.~\ref{eq:act} has only one minimum
    (it is convex), and can be easily solved using standard methods
    such as employed in our previous work\cite{}.
We employ such an alternating directions strategy,
    where both $H_z$ and $\epsilon^{-1}$ are solved independently.
This process is extremely inefficient,
    but is employed because the underlying numerical methods
    do not require any tuning.
We expect considerable improvements in computational efficiency
    when more sophisticated algorithms are applied,
    especially those which can update $H_z$ and $\epsilon^{-1}$ independently.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{quickfig/fiber} 
    \caption{Waveguide coupler for a wide, low-index 
        ($\epsilon=2.25$) waveguide. 
        The dielectric structure of the coupler and surrounding input and
        output waveguides is shown on the left, while the simulation
        validating our results is shown on the right.
        The coupler converts $96.3\%$ of the input power to the
        designated output mode.
        The device is extremely compact, 
        convering only $36 \times 66$ grid points,
        where the vacuum wavelength is 42 grid points.
        Computation time was 20 minutes on a personal computer.}
    \label{fig:fiber}
\end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{quickfig/mode-conv} 
    \caption{Coupler that converts the fundamental waveguide mode to the
        second-order waveguide mode.
        This problem is quite difficult since the two modes are of 
        opposite symmetry.
        For example, adiabatic approaches cannot be applied to this case.
        However, our method produces a device 
        (which has the same dimensions and vacuum wavelength as Fig. 1) 
        which achieves a coupling efficiency of $95.5\%$. 
        Computation time was 20 minutes on a personal computer. (movie?)
        }
    \label{fig:mode}
    \end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{quickfig/air-core}
    \caption{Coupler to an air-core mode.
        Here, not only are the modes of opposite symmetry,
        but the output waveguide operates on a fundamentally different
        principle (guided by Bragg reflection) than the input waveguide 
        (index guided).
        The device still achieves an efficiency of $83.3\%$, demonstrating the
        versatility of our method.
        The vacuum wavelength is 25 grid points, 
        while the device footprint is still $36 \times 66$ grid points.
        Computation time was 20 minutes on a personal computer.
        }
        \label{fig:aircore}
\end{figure}

% \section{Conclusion}
% We introduce an objective-first approach to structure design,
%     and apply it to the design of nanophotonic waveguide couplers.
% We show that our method is
%     computationally fast,
%     generates high-efficiency devices with compact footprints,
%     does not require a good starting design, and
%     can be applied to arbitrary waveguide modes.
% 
%

This work has been supported in part by the 
    AFOSR MURI for Complex and Robust On-chip Nanophotonics 
    (Dr. Gernot Pomrenke), grant number FA9550-09-1-0704.

>>>>

\begin{document}
\title{Objective-First Design of Nanophotonic Waveguide Couplers}
\author{Jesse Lu and Jelena Vu\v{c}kovi\'{c}}
\address{Stanford University, Stanford, California, USA.}
\email{jesselu@stanford.edu}

\maketitle
\begin{abstract}
We introduce an ``objective-first'' approach to the design of
    nanophotonic waveguide couplers.
Our method is
    computationally fast (20 minutes), 
    requires no trial-and-error,
    does not require a good starting design,
    can be applied to arbitrary waveguide modes,
    and generates devices with high coupling efficiencies (typically 95\%)
    and small footprints (1-4 square vacuum wavelengths).
\end{abstract}
\ocis{230.7370, 130.3990.}

\begin{thebibliography}{99}
\bibitem{fibergrating} Y. Tang, Z. Wang, L. Wosinski, U. Westergren, and S. He,
    ``Highly efficient nonuniform grating coupler for silicon-on-insulator 
    nanophotonic circuits,''
    Opt. Lett. \textbf{35}, 1290-1292 (2010)
\bibitem{ridge}  K. K. Lee, D. R. Lim, L.C. Kimerling, J. Shin, and F. Cerrina, 
    ``Fabrication of ultralow-loss Si/SiO2 waveguides by roughness reduction,''
    Opt. Lett. \textbf{26}, 1888-1890 (2001)
\bibitem{pcslow} Y. A. Vlasov, M. O'Boyle, H. F. Hamann, and S. J. McNab,
    ``Active control of slow light on a chip with photonic crystal waveguides,''
    Nature \textbf{438}, 65-69 (2005)
\bibitem{slotfocus} M. Lipson, 
    ``Guiding, Modulating, and Emitting Light on 
    Silicon-Challenges and Opportunities,'' 
    J. Lightwave Technol. \textbf{23}, 4222-4238 (2005) 
\bibitem{active} J. Van Campenhout, P. Rojo Romeo, P. Regreny, C. Seassal, 
    D. Van Thourhout, S. Verstuyft, L. Di Cioccio, J.-M. Fedeli, 
    C. Lagahe, and R. Baets, 
    ``Electrically pumped InP-based microdisk lasers integrated with a 
    nanophotonic silicon-on-insulator waveguide circuit,'' 
    Opt. Express \textbf{15}, 6744-6749 (2007) 
\bibitem{metallic} L. Tang, S. E. Kocabas, S. Latif, A. K. Okyay, 
    D. S. Ly-Gagnon, K. C. Saraswat and D. A. B. Miller, 
    ``Nanometre-Scale Germanium Photodetector Enhanced by a 
    Near-Infrared Dipole Antenna,'' 
    Nature Photonics \textbf{2}, 226 – 229 (2008) 
\bibitem{fwadia} V. R. Almeida, R. R. Panepucci, and M. Lipson, 
    ``Nanotaper for compact mode conversion,'' 
    Opt. Lett. \textbf{28}, 1302-1304 (2003) 
\bibitem{wwadia} S. G. Johnson, P. Bienstman,  M. A. Skorobogatiy, 
    M. Ibanescu1, E. Lidorikis, and J. D. Joannopoulos,
    ``Adiabatic theorem and continuous coupled-mode theory for 
    efficient taper transitions in photonic crystals,''
    Phys. Rev. E \textbf{66}, 066608 (2002)
\bibitem{deriv} F. Wang, J. S. Jensen, O. Sigmund, 
    ``Robust topology optimization of photonic crystal waveguides with 
    tailored dispersion properties.'' 
    J. Opt. Soc. Am. B \textbf{28}, 387-397 (2011)
\bibitem{boydbook} S. Boyd, and L. Vandenberghe, 
    \emph{Convex Optimization} 
    (Cambridge University Press, 2004)
\bibitem{prevwork} J. Lu, S. Boyd, and J. Vuckovic, 
    ``Inverse design of a three-dimensional nanophotonic resonator,''
    Opt. Express \textbf{19}, 10563-10570 (2011) 
\bibitem{code} \url{www.github.com/JesseLu/objective-first}
\end{thebibliography}

@@

% Why is waveguide mode conversion important?
\subsection{The importance of waveguide mode conversion}
Optical mode conversion, 
    the efficient transfer of photons from one guided mode to another,
    is a fundamental requirement in nanophotonics.
Efficient conversion between waveguides modes
    is critical in the many cases, including:
    \begin{enumerate}
        \item Coupling to and from optical fiber\cite{fibergrating}, 
        to communicate with the outside world.
    \item Coupling between various nanophotonic waveguides, 
        since different waveguides are best suited for different applications.
        For example, ridge waveguides seem ideal for 
            low-loss transport\cite{ridge},
            but other waveguides, 
            such as photonic crystal waveguides or slot waveguides,
            may be better suited for slow-light\cite{pcslow} 
            or energy-focusing devices\cite{slotfocus}.
    \item Coupling between different materials systems such as
       passive, active\cite{active}, and metallic\cite{metallic} devices. 
    \end{enumerate}
Although efficient waveguide coupling is essential in any nanophotonic system,
    practical methods to design couplers have not yet been developed.

% Why are current solutions insufficient?
\subsection{Approaches to designing waveguide couplers}
% Brute force
Of the design strategies currently available,
    brute-force parameter search is the most often employed 
    because of its sheer simplicity.
Although it may be suitable for tuning existing designs,
    the parameter space for most practical devices is simply too large
    for such a strategy to be tractable.

% Adiabatic mode conversion (large devices, symmetry breaking)
Adiabatic mode conversion strategies have been succesful
    for certain fiber-waveguide\cite{fwadia} and 
    waveguide-waveguide\cite{wwadia} couplers,
    even if the resulting devices often require large footprints.
However, adiabatic approaches cannot be used in many important cases 
    such as coupling in the out-of-plane direction, or
    coupling between modes of opposite symmetry.

% Local optimization
On the other hand, optimization methods based on local topological derivatives 
    seem very promising\cite{deriv}
    in that they are both much faster than brute-force methods and
    more adaptable than adiabatic strategies.
However, these methods still carry a significant computational burden 
    in that every updated design must be simulated at least once.
Also, there is a significant burden for the user from whom
    a good initial design is usually required.

% How does your algorithm solve this problem?
% \subsection{Advantages of objective-first design}
In contrast, our method
\begin{itemize}
    \item does not employ brute-force parameter searches,
    \item does not require a good initial design,
    \item is computationally fast (no simulations required),
    \item generates couplers between seemingly arbitrary waveguide modes, and
    \item can generate these couplers within a very small footprint. 
\end{itemize}
We outline the general design strategy,
    as well as how it is applied to coupler design, in the section below.

@@

The differences between Eqs.~\ref{eq:adj} and \ref{eq:ob1} are:
\begin{enumerate}
    \item in Eq.~\ref{eq:adj:obj}, we attempt to decrease the design objective,
        while in Eq.~\ref{eq:ob1:con}, the design objective is kept at zero;
    \item in Eq.~\ref{eq:adj:con}, we always satisfy the underlying physics,
        while in Eq.~\ref{eq:ob1:obj}, physics is not satisfied,
            since the physics residual is generally non-zero.
\end{enumerate}

@@

The first practical implication of the objective-first approach 
    is that the number of independent variables is increased to include
    both $x$ and $p$.
This occurs because in Eq.~\ref{eq:adj}, 
    the constraint that physics must be satisfied, $g(x,p)=0$, 
    essentially forces $x$ to be a dependentent variable,
    since the choice of $p$ implicitly determines the value of $x$
    (there is generally a one-to-one mapping from $p$ to $x$).
In contrast, Eq.~\ref{eq:ob1} allows both $x$ and $p$ to vary independently,
    because the constraint, $f(x)=0$, is only a function of $x$.

Secondly, the amount of computation needed to enforce the constraint is
    drastically reduced in the objective-first approach.
This is because Eq.~\ref{eq:adj} requires a full solution of $g(x,p)=0$
    (i.e. a full simulation of the structure, $p$) to compute $x$.
In contrast, the constraint $f(x)=0$ in Eq.~\ref{eq:ob1} 
    can often be enforced with a trivial amount of computation (as shown below).

Lastly, the objective-first approach eliminates the need for even a reasonable
    initial design.
Generally, methods based on Eq.~\ref{eq:adj} require an initial design which
    already provides some limited functionality
    (e.g. a coupler which already transfers 
    a non-zero amount of power to the desired output mode).
In constrast, out method, based on Eq.~\ref{eq:ob1}, performs just as well
    when started from a completely non-functional design 
    (e.g. a coupler which transfers no power into the desired output mode).

% How is it applied to waveguide coupler design?
\subsection{Objective-first approach to waveguide coupler design}
We now apply the objective-first approach to the problem of designing 
    two-dimensional nanophotonic waveguide couplers.

We choose to work in the two-dimensional transverse electric mode,
    which only couples $E_x$, $E_y$, and $H_z$ ($E_z, H_x, H_y = 0$),
    since it is most relevant for on-chip devices.
We choose to use $H_z$ as the field variable, 
    and $\epsilon^{-1}$ (inverse of the permittivity) 
    as the structure variable.
This results in the following representation of the physics residual, 
    based on the time-harmonic electromagnetic wave equation without sources;
    \begin{equation}
    \|g(H_z, \epsilon^{-1})\|^2 = 
    \| \nabla \times \epsilon^{-1} \nabla \times H_z - \mu \omega^2 H_z \|^2,
    \end{equation}
    where $\omega$ is the angular frequency,
    and $\mu$ is the permeability of free-space.

@@

Finally, as in any method vase on an objective-first approach,
    the physics residual is not guaranteed to decrease to zero.
Thus, it is entirely possible to never achieve 
    a physically realizable field, $H_z$.
In such cases, which are the norm rather than the exception,
    we find that a relatively small residual usually leads to
    fairly good, although imperfect, device performance.

\subsection{Numerical methods used to solve the objective-first design problem}
The design problem is now
    \begin{subequations}\label{eq:act}
    \begin{align} 
    \text{decrease} & \quad  
        \| \nabla \times \epsilon^{-1} \nabla \times H_z - \mu \omega^2 H_z \|^2
        \label{eq:act:obj} \\ 
    \text{subject to} & \quad 
        \begin{bmatrix}
        H_z - H_z^\text{perfect} \\
        \frac{\partial H_z}{\partial n} - 
            \frac{\partial H_z^\text{perfect}}{\partial n}
        \end{bmatrix}_\text{boundary}
        = 0.
    \end{align}
    \end{subequations}

This problem contains many local minima (it is non-convex\cite{boydbook});
    however, when either the field ($H_z$) or the structure ($\epsilon^{-1}$)
    variable is considered separately Eq.~\ref{eq:act} has only one minimum
    (it is convex), and can be easily solved using standard methods
    such as employed in our previous work\cite{prevwork}.
We employ such an alternating directions strategy,
    where both $H_z$ and $\epsilon^{-1}$ are solved independently.
This process is extremely inefficient,
    but is employed because the underlying numerical methods
    do not need to be tuned by the user.
We expect considerable improvements in computational efficiency
    when more sophisticated algorithms are applied,
    especially those which can update $H_z$ and $\epsilon^{-1}$ independently.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{quickfig/fiber} 
    \caption{Waveguide coupler for a wide, low-index 
        ($\epsilon=2.25$) waveguide. 
        The dielectric structure of the coupler and surrounding input and
        output waveguides is shown on the left, while the simulation
        validating our results is shown on the right.
        The coupler converts $96.3\%$ of the input power to the
        designated output mode.
        The device is extremely compact, 
        convering only $36 \times 66$ grid points,
        where the vacuum wavelength is 42 grid points.
        Computation time was 20 minutes on a personal computer.}
    \label{fig:fiber}
\end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{quickfig/mode-conv} 
    \caption{Coupler that converts the fundamental waveguide mode to the
        second-order waveguide mode
        (a movie of the design progress is included as a
        supplementary material).
        This problem is quite difficult since the two modes are of 
        opposite symmetry.
        For example, adiabatic approaches cannot be applied to this case.
        However, our method produces a device 
        (which has the same dimensions and vacuum wavelength as Fig. 1) 
        with a coupling efficiency of $95.5\%$. 
        Computation time was 20 minutes on a personal computer.
        }
    \label{fig:mode}
    \end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{quickfig/air-core}
    \caption{Coupler to an air-core mode.
        Here, not only are the modes of opposite symmetry,
        but the output waveguide operates on a fundamentally different
        principle (distributed reflection) than the input waveguide 
        (index guided).
        The device still achieves an efficiency of $83.3\%$, demonstrating the
        versatility of our method.
        The vacuum wavelength is 25 grid points, 
        while the device footprint is still $36 \times 66$ grid points.
        Computation time was 20 minutes on a personal computer.
        }
        \label{fig:aircore}
\end{figure}

This work has been supported in part by the 
    AFOSR MURI for Complex and Robust On-chip Nanophotonics 
    (Dr. Gernot Pomrenke), grant number FA9550-09-1-0704.
The Matlab code used is available online\cite{code}.
